---
title: "chapter6"
author: "Sadiksha Adhikari"
date: "2022-12-08"
output: html_document
---

```{r}
date()
```

# ASSIGNMENT 6: ANALYSIS OF LONGITUDINAL DATA

## Data wrangling
Data wrangling part completed and saved by the name meet_and_repeat.R in data folder.

## Data analysis

After all these weeks of hard work, you should be able to go (almost) on your own. Please use the corresponding Exercise Set and the MABS4IODS materials: It should not be too difficult. Instead, it should be interesting and fun! (You should see NEW RESULTS that do NOT APPEAR in MABS!)
Note: As you read the data sets in your chapter6.Rmd you may have to factor the categorical variables again, as R may not recognise them automatically as factors.

Implement the analyses of Chapter 8 of MABS, using the R codes of Exercise Set 6: Meet and Repeat: PART I but using the RATS data (from Chapter 9 and Meet and Repeat: PART II).
(0-7 points: 0-4 points for graphs or analysis results + 0-3 points for their interpretations)

## PART I on RATS data
## Load and explore RATS data
```{r }
# read in the RATS data
RATSL <- read.table("data/RATSL.csv", header = TRUE, sep = ',')

library(dplyr)
# Factor variables ID and Group
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

# Look at the data
glimpse(RATSL)
```

This is the long format of rats dataset. It has 176 rows and 5 columns. It contains the columns for ID, Group, WD, Weight and Time. ID and Group variables are factor types,WD is character, Weigh and Time are integers.

## Visualization
```{r }
#Access the package ggplot2
library(ggplot2)
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)") +
  scale_y_continuous(name = "Weight (grams)")+
  theme(legend.position = "top")

```

## Standarise variable weight and visualization

```{r }
# Standardise the variable Weight

RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate( stdWeight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)

# Plot again with the standardised Weight
library(ggplot2)
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) + #edit here on how many
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized Weight")

```

Rats in groups 2 and 3 gain more weight with increasing weeks. This observation is clearer with standardised data. Group 1 rats do not seem to have gained as much weight compared to the rest.


## Mean and standard error profiles of Weight by Group and Time and visualizaton

```{r }
# Number of IDs (per group):
sum(RATSL$Group == 1)
sum(RATSL$Group == 2)
sum(RATSL$Group == 3)
```


```{r }

# Number of IDs (per group):
IDs_per_group_count <- RATSL %>% group_by(ID, Group) %>%summarise( count = n())
n <- 44

library(dplyr)
library(tidyr)

# Summary data with mean and standard error of Weight by Group and Time 
WeightS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = Weight, se = Weight ) %>%
  ungroup()

# Glimpse the data
glimpse(WeightS)

# Plot the mean profiles
library(ggplot2)
ggplot(WeightS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  #geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")

# Plot the mean profiles with geom_errorbar
library(ggplot2)
ggplot(WeightS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")




```

This shows that rats are slightly gaining more weight over time. There is some overlap in the mean profiles of the group 2 and 3, at least with two values. So there might probably be non-significant difference between the two groups in respect to the mean weight. But group 1 mean does not overlap at all suggesting high difference in mean weight of this group compared to group 2 and 3.

## Finding the outlier

```{r }
library(dplyr)
library(tidyr)
# Create a summary data by Group and ID with mean as the summary variable (ignoring baseline Time 0)
RATSL8S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL8S)

# Draw a boxplot of the mean versus Group
library(ggplot2)
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), 9 weeks period")

```

## Dataset without outlier

Based on the plot above it would make sense to deal with the outliers separately for each group. 

```{r }
Rats1 <- RATSL8S[RATSL8S$Group == 1,]
Rats2 <- RATSL8S[RATSL8S$Group == 2,]
Rats3 <- RATSL8S[RATSL8S$Group == 3,]

group1_Q1 <- quantile(Rats1$mean, .25)
group1_Q3 <- quantile(Rats1$mean, .75)
group1_IQR <- IQR(Rats1$mean)

group2_Q1 <- quantile(Rats2$mean, .25)
group2_Q3 <- quantile(Rats2$mean, .75)
group2_IQR <- IQR(Rats2$mean)

group3_Q1 <- quantile(Rats3$mean, .25)
group3_Q3 <- quantile(Rats3$mean, .75)
group3_IQR <- IQR(Rats3$mean)

Rats1 <- subset(Rats1, Rats1$mean > (group1_Q1 - 1.5*group1_IQR) & Rats1$mean < (group1_Q3 + 1.5*group1_IQR))
Rats2 <- subset(Rats2, Rats2$mean > (group2_Q1 - 1.5*group2_IQR) & Rats2$mean < (group2_Q3 + 1.5*group2_IQR))
Rats3 <- subset(Rats3, Rats3$mean > (group3_Q1 - 1.5*group3_IQR) & Rats3$mean < (group3_Q3 + 1.5*group3_IQR))

RATSL8S_backup <- RATSL8S #just to check and keep backup in case if I need to re-do something (if I mess up with this data) 

RATSL8S <- rbind(Rats1, Rats2, Rats3)

# Glimpse the data
glimpse(RATSL8S)

# Draw a boxplot of the mean versus Group
library(ggplot2)
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), 9 weeks period") 

```

We see that the mean values have been corrected with the removal of outliers. The values that were overlapping between group 2 and 3 were actually outliers. So the mean between group 2 and 3 are not that close as we previously thought. 

## Pairwise comparisons

```{r }
# Pairwise comparisons instead of two sample t test because we have 3 groups
library(rstatix)
pwc <- RATSL8S %>%
  pairwise_t_test(mean ~ Group, p.adjust.method = "bonferroni")
pwc

library(dplyr)
library(tidyr)
library(ggpubr)

# Show significance levels
library(ggpubr)

ggplot(RATSL8S, aes(x = Group, y = mean)) +
  stat_pvalue_manual(pwc, hide.ns = TRUE, label = "p.adj.signif", y.position = c(475,550,540))+
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), 9 weeks period")
```

From pairwise comparison, we see that the mean weight for all of the groups are different from one another. It is also visible from the plot

## Anova test
```{r }
# Add the baseline from the original data as a new variable to the summary data
# For that we need to remove outlers also from RATS data
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')
RATS_group_id <- paste0(RATS$ID, "_", RATS$Group)
RATSL8S_group_id <- paste0(RATSL8S$ID, "_", RATSL8S$Group) 

setdiff(RATS_group_id,RATSL8S_group_id)

```

```{r }
#manually removing those outliers from rats data
RATS <-  RATS[-c(2,12,13),]

RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)

#let's also try this 
res.aov <- RATSL8S2 %>% anova_test(mean ~ baseline+ Group) 
res.aov


# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline+ Group, data = RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)
```

Also according to anova test there is significant difference in weight change between the groups and from the baseline.

## PART II on BPRS data, linear mized effects model

## Load the data and visualize

```{r }

# read in the BPRS data
BPRSL <- read.table("data/BPRSL.csv", header = TRUE, sep = ',')

library(dplyr)
# Factor variables subject and treatment
BPRSL$subject <- factor(BPRSL$subject)
BPRSL$treatment <- factor(BPRSL$treatment)

# Glimpse the data
glimpse(BPRSL)

# Plot the BPRSL data
library(ggplot2)
ggplot(BPRSL, aes(x = week, y = bprs, treatment = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Weeks") +
  scale_y_continuous(name = "bprs")+
  theme(legend.position = "top")


```

With this plot it is not possible to say if there is any difference in results from the treatment 

## The Random Intercept Model
```{r }
# create a regression model BPRS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)

summary(BPRS_reg)

# Create a random intercept model
library(lme4)
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)


```

We see significant difference in weeks taken.

## Random Intercept and Random Slope Model

```{r }


# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)

```


## weeks to interact: Random Intercept and Random Slope Model with interaction

```{r }

# create a random intercept and random slope model with the interaction
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | treatment), data = BPRSL, REML = FALSE)

summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

library(dplyr)
library(tidyr)

# Create a new column fitted to BPRSL
BPRSL <- BPRSL %>% mutate(fitted = Fitted)

# draw the plot of BPRSL with the Fitted values of bprs
library(ggplot2)
ggplot(BPRSL, aes(x = week, y = Fitted, treatment = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)") +
  scale_y_continuous(name = "Fitted bprs") +
  theme(legend.position = "top")
```

It appears that there is no significant difference in results between the two treatment groups based on the tests, also visible from the plot.

```{r }

```

```{r }

```

```{r }

```